---
title: "Concentration-response analysis"
params:
  manifest: ""
  dev_package_dir: null
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    geometry: margin=2.5cm
    fig-pos: "H"
    include-in-header:
      text: |
        \usepackage{float}
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "ragg_png"
    dpi: 200
---

```{r setup}
#| include: false
options(width = 88)

## toxtools is loaded by name when installed, and from source when the report
## is being rendered from a development copy of the package. run_workbook()
## decides which and passes the source directory through when it applies.
if (!is.null(params$dev_package_dir) && nzchar(params$dev_package_dir)) {
  pkgload::load_all(params$dev_package_dir, quiet = TRUE)
} else {
  library(toxtools)
}
library(bayesnec)
library(ggplot2)

m <- readRDS(params$manifest)

## Sheet names are whatever the laboratory typed. In a PDF they are set by
## LaTeX, where several ordinary characters are commands, so they are escaped
## before being written into a heading.
esc <- function(x) {
  if (!knitr::is_latex_output()) {
    return(x)
  }
  x <- gsub("\\\\", "\\\\textbackslash{}", x)
  gsub("([&%$#_{}])", "\\\\\\1", x)
}

fmt <- function(x, digits = 3) {
  formatC(signif(x, digits), format = "g", digits = digits, drop0trailing = TRUE)
}

## Names a set of equations and the weight they hold between them. Written as a
## function because the number of equations is not known in advance and the
## sentence has to read correctly for one as well as for several.
carrying <- function(models, weight) {
  paste0(
    paste(models, collapse = ", "),
    if (length(models) > 1) ", which between them carry " else ", which carries ",
    sprintf("%.2f", 100 * sum(weight, na.rm = TRUE)),
    " per cent of the weight in the model average"
  )
}
```

Workbook: **`r esc(m$workbook)`**

Analysed `r format(m$created, "%d %B %Y")` with `bayesnec`
`r m$bayesnec_version`, fitting the **`r m$model_set`** model set to
`r m$predictor_transform`-transformed concentration.

# What was analysed

Each sheet of the workbook was examined for a row of column names giving a
concentration column and a response column, with numeric data beneath it.
Sheets holding such a block were analysed; the rest were skipped, for the
reasons given below.

```{r classification}
cls <- m$classification
## `analyse` is whether the sheet could be fitted; `analysed` is whether it
## was. They differ when the run named a subset of sheets.
shown <- if ("analysed" %in% names(cls)) cls$analysed else cls$analyse
cls$analysed <- ifelse(shown, "yes", "no")
cls$reason[is.na(cls$reason)] <- ""
knitr::kable(
  cls[, c("sheet", "analysed", "header_row", "n_rows", "n_levels", "response",
          "reason")],
  col.names = c("Sheet", "Analysed", "Names on row", "Rows",
                "Concentrations", "Response", "Reason skipped"),
  booktabs = TRUE, longtable = TRUE
)
```

## How to read this report

For each analysed sheet the report gives, in order: the data as they were read
from the sheet; the equations that were fitted and the weight each carries in
the model average; the no-effect and effect concentrations; the fitted curve;
the individual equations behind the average; and the diagnostic checks.

The **N(S)EC** is the model-averaged no-effect concentration. It combines the
`nec` parameter of the threshold equations with the no-significant-effect
concentration of the smooth ones. **EC10** and **EC50** are the concentrations
at which the response has declined by 10 and 50 per cent from the fitted
control value. All three are reported in the units recorded on the sheet, with
a 95 per cent credible interval.

The diagnostics decide whether those numbers can be relied on. Look for three
things, in each case **in an equation that carries weight**: an R-hat above
1.05, which means the sampler did not converge; over-dispersion, which means
the counts vary more than a binomial response allows and the intervals are
narrower than they should be; and chain traces that do not overlap or that
drift, which mean the same as a high R-hat.

The weight qualification matters, and it is why both checks are reported
against the weight each equation carries. Every equation in the set is fitted,
including ones whose shape suits the data badly. Such an equation fails the
dispersion test whatever the data are doing, because the variation its curve
cannot account for is counted as dispersion, and it is often the one the
sampler struggles with as well. It is also given almost no weight in the model
average, for the same reason. So a check it fails says something about that
curve and nothing about the estimates reported here. Both tables below are
ordered by weight, and the sentence above each warns only about equations
carrying at least one per cent of it.

```{r sections}
#| output: asis
child_src <- c(
  "# `r esc(entry$sheet)`",
  "",
  "```{r}",
  "#| include: false",
  "obj <- readRDS(entry$fit_file)",
  "sheet <- obj$sheet",
  "est <- entry$estimates",
  "diag <- workbook_diagnostics(obj)",
  "## With no transformation the printed summary is already in the recorded",
  "## units, and warning about a difference that does not exist would only",
  "## confuse the reader.",
  "scale_note <- if (obj$transform == 'identity') {",
  "  'The estimates below are in the units recorded on the sheet.'",
  "} else {",
  "  paste0('The N(S)EC printed here is on the ', obj$funs$note, ' scale ',",
  "         'the curve was fitted on, which is what the sampler worked ',",
  "         'with. The next section gives the same quantity in the units ',",
  "         'recorded on the sheet, so the two numbers differ.')",
  "}",
  "```",
  "",
  "## Data",
  "",
  "Read from sheet `r esc(sheet$sheet)`, column names on row",
  "`r sheet$header_row`. Response: `r sheet$response`, fitted with",
  "`r sheet$family_call`.",
  "",
  "```{r}",
  "d <- sheet$data",
  "if (!is.na(sheet$trials_var)) {",
  "  d$proportion <- d[[sheet$y_var]] / d[[sheet$trials_var]]",
  "}",
  "agg <- do.call(rbind, lapply(split(d, d$concentration), function(g) {",
  "  yv <- if ('proportion' %in% names(g)) g$proportion else g[[sheet$y_var]]",
  "  data.frame(",
  "    concentration = g$concentration[1], n = nrow(g),",
  "    mean = mean(yv), min = min(yv), max = max(yv)",
  "  )",
  "}))",
  "agg <- agg[order(agg$concentration), ]",
  "knitr::kable(",
  "  agg, row.names = FALSE, digits = 4, booktabs = TRUE,",
  "  col.names = c('Concentration', 'Replicates', 'Mean', 'Minimum',",
  "                'Maximum'),",
  "  caption = 'Recorded response by concentration.'",
  ")",
  "```",
  "",
  "```{r}",
  "#| fig-height: 3.2",
  "#| fig-cap: 'Data as read from the sheet, before any model was fitted.'",
  "plot_sheet_data(sheet, transform = obj$transform)",
  "```",
  "",
  "`r if (length(sheet$notes)) paste0('Note: ', sheet$notes, collapse = ' ')`",
  "",
  "## Fitted model set",
  "",
  "The equations retained, the weight each carries in the model average, and",
  "the fit of each. `r scale_note`",
  "",
  "```{r}",
  "#| comment: ''",
  "summary(obj$fit)",
  "```",
  "",
  "## No-effect and effect concentrations",
  "",
  "```{r}",
  "tab <- data.frame(",
  "  Estimate = est$estimate,",
  "  Value = fmt(est$value),",
  "  Lower = fmt(est$lower),",
  "  Upper = fmt(est$upper)",
  ")",
  "knitr::kable(",
  "  tab, row.names = FALSE, booktabs = TRUE,",
  "  col.names = c('Estimate', 'Median', 'Lower 95%', 'Upper 95%'),",
  "  caption = paste0('Model-averaged estimates, in ',",
  "                   sheet$labels$x, '.')",
  ")",
  "```",
  "",
  "## Model-averaged fit",
  "",
  "```{r}",
  "#| fig-height: 4",
  "#| fig-cap: 'Model-averaged curve with its 95% credible band. The vertical",
  "#|   lines mark the N(S)EC and its credible interval.'",
  "plot_workbook_fit(obj, estimates = est, title = NULL)",
  "```",
  "",
  "## Individual equations",
  "",
  "```{r}",
  "#| include: false",
  "## The panels are laid out in two columns, so the figure has to grow with",
  "## the number of equations. Left at a fixed height, a set of three would",
  "## be stretched and a set of twelve would overflow the page, which with",
  "## figures held in place rather than floated leaves a blank page behind.",
  "fig_h <- min(8.4, 2.1 * ceiling(length(obj$fit$success_models) / 2) + 0.4)",
  "```",
  "",
  "```{r}",
  "#| fig-height: !expr fig_h",
  "#| fig-cap: 'Each equation retained in the model set, fitted separately.",
  "#|   The averaged curve above is a weighted combination of these.'",
  "plot_workbook_fit(obj, all_models = TRUE, title = NULL)",
  "```",
  "",
  "## Diagnostics",
  "",
  "```{r}",
  "#| output: asis",
  "rt <- diag$rhat_flags",
  "hit <- if (is.null(rt)) rt[0, ] else rt[rt$flagged, , drop = FALSE]",
  "if (is.null(rt) || nrow(hit) == 0) {",
  "  cat('No equation was flagged for non-convergence; every R-hat is at or ',",
  "      'below 1.05.', sep = '')",
  "} else if (any(hit$warn)) {",
  "  bad <- hit[hit$warn, , drop = FALSE]",
  "  cat('**The sampler did not converge** for ',",
  "      paste(sprintf('%s (weight %.2f)', bad$model, bad$weight),",
  "            collapse = ', '),",
  "      '. These equations carry weight in the model average, so the ',",
  "      'estimates above should not be relied on until they are refitted ',",
  "      'with more iterations.', sep = '')",
  "} else {",
  "  cat('The sampler did not converge for ',",
  "      carrying(hit$model, hit$weight),",
  "      ', so the estimates above are unaffected. An equation the sampler ',",
  "      'struggles with is usually one that suits the data badly, which is ',",
  "      'also why it is given no weight. Every other R-hat is at or below ',",
  "      '1.05.', sep = '')",
  "}",
  "```",
  "",
  "```{r}",
  "r2_names <- c('Model', 'R-squared', 'Standard error', 'Lower 95%',",
  "              'Upper 95%')",
  "if (!is.null(diag$r2)) {",
  "  knitr::kable(",
  "    diag$r2, row.names = FALSE, digits = 3, booktabs = TRUE,",
  "    col.names = if (ncol(diag$r2) == length(r2_names)) {",
  "      r2_names",
  "    } else {",
  "      colnames(diag$r2)",
  "    },",
  "    caption = 'Bayesian R-squared for each equation.'",
  "  )",
  "}",
  "```",
  "",
  "```{r}",
  "#| output: asis",
  "dd <- diag$dispersion",
  "if (is.null(dd)) {",
  "  cat('The over-dispersion test does not apply to this response type.')",
  "} else {",
  "  warn_rows <- dd[dd$warn, , drop = FALSE]",
  "  quiet_rows <- dd[dd$overdispersed & !dd$warn, , drop = FALSE]",
  "  if (nrow(warn_rows)) {",
  "    cat('**Over-dispersion is present.** ',",
  "        paste(sprintf('%s (weight %.2f, dispersion %.2f, %.2f-%.2f)',",
  "                      warn_rows$model, warn_rows$weight,",
  "                      warn_rows$estimate, warn_rows$lower,",
  "                      warn_rows$upper), collapse = '; '),",
  "        '. These equations carry weight in the model average, so the ',",
  "        'counts vary more than a binomial response allows and the ',",
  "        'credible intervals reported above are narrower than they ',",
  "        'should be. Refitting with a beta-binomial response is ',",
  "        'warranted.', sep = '')",
  "  } else if (nrow(quiet_rows)) {",
  "    cat('No equation that carries weight in the model average is ',",
  "        'over-dispersed. The test is failed by ',",
  "        carrying(quiet_rows$model, quiet_rows$weight),",
  "        '. An equation that fits badly fails this test because of its fit ',",
  "        'rather than because of the data, so this is not a reason to ',",
  "        'refit.', sep = '')",
  "  } else {",
  "    cat('No equation shows over-dispersion.')",
  "  }",
  "  cat('\\n\\n')",
  "  show <- data.frame(",
  "    dd$model, round(dd$weight, 3), round(dd$estimate, 2),",
  "    round(dd$lower, 2), round(dd$upper, 2),",
  "    ifelse(dd$overdispersed, 'yes', 'no'),",
  "    ifelse(dd$warn, 'yes', 'no')",
  "  )",
  "  print(knitr::kable(",
  "    show, row.names = FALSE, booktabs = TRUE,",
  "    col.names = c('Model', 'Weight', 'Dispersion', 'Lower 95%',",
  "                  'Upper 95%', 'Over-dispersed', 'Worth acting on'),",
  "    caption = paste('Over-dispersion test, ordered by the weight each',",
  "                    'equation carries in the model average. An interval',",
  "                    'excluding 1 fails the test; it is worth acting on',",
  "                    'only where the equation also carries weight.')",
  "  ))",
  "}",
  "```",
  "",
  "### Chain traces",
  "",
  "One page per equation. The left column is the autocorrelation of each",
  "chain, the right column the sampled values. Chains that overlap and show",
  "no trend indicate the sampler converged. The parameters are on the",
  "`r obj$funs$note` scale the curve was fitted on, so a `nec` trace is not",
  "in the units recorded on the sheet.",
  "",
  "```{r}",
  "#| fig-height: 7",
  "#| fig-width: 6.5",
  "check_chains(obj$fit)",
  "```",
  ""
)

for (entry in m$entries) {
  cat(knitr::knit_child(text = child_src, envir = environment(),
                        quiet = TRUE), sep = "\n")
  cat("\n\n")
}
```

# Method

Curves were fitted with `bayesnec::bnec()`, which fits every equation in the
requested model set and combines them by stacking weights into a single
model-averaged prediction. The `r m$model_set` set was used. Equations that
cannot be fitted to the response and link in use are dropped by `bnec()` before
sampling; the equations that were retained are listed for each sheet above.

Concentration was `r m$predictor_transform`-transformed before fitting. For a
dilution series that includes a zero control, a square root is used rather than
a logarithm, because the logarithm of zero is undefined and a log-scaled fit
would either drop the controls or require an arbitrary amount added to every
concentration. Estimates are back-transformed to the recorded units before
reporting.

Every family is fitted with an identity link, so that the fitted `top`, `bot`
and `nec` parameters are on the scale of the response itself and can be read
directly. Counts are fitted as counts rather than as a per cent of control:
`bayesnec::ecx()` measures the decline against the fitted control value one
posterior draw at a time, which keeps the uncertainty in the control, whereas
dividing by the observed control mean beforehand discards it.

```{r}
#| output: asis
## No esc() here: the path is written inside a code span, which pandoc
## escapes on its own. Escaping first leaves the escape visible, so a path
## holding an underscore renders as "Test\\_data".
cat("Fits are saved in `", file.path(m$run_dir, "fits"),
    "` and can be reloaded with `readRDS()`.", sep = "")
```
